# Lab 1: Succession {#lab1}

Some guidelines 

## Results

You should provide data and analyses to address the following questions:

 - What are the qualitative trends and characteristics of each habitat?
 - How does the total density of canopy trees differ between habitats?
 - Does relative abundance of each species differ between canopy and sapling trees?  How consistent is this among habitats?

### Canopy tree density

You can use the point-quarter method to estimate tree density at each site with a bit of math.  Let $x_i$ be the distance from your point to a sampled canopy tree. If you sampled 4 trees at a point, then the density in trees per square meters would be $\frac{4}{x_1^2 + x_2^2 + x_3^2 + x_4^2}$. More generally, the density is $1 / \text{mean}(x^2)$.  

Calculate density for each sample point, and convert it to hectares (multiply by $10,000 \text{ m}^2/\text{ha}$).  How do densities compare between habitat types?  Create a figure and compare the averages. Note that you won't be able to claim that one group is different unless you use statistical tests, such as a one-way ANOVA (this is optional).

### Species by habitat and life stage

For each habitat, you will want to calculate the relative abundance of each tree species for canopy and sapling trees (i.e., number of Canopy species *x* in habitat *y* divided by total number of observed Canopy trees in *y*). Visualize the result for each habitat (a three-panel frequency plot would be the best way to go, with categories ordered by canopy abundance). 

To test if the relative proportions in canopy and saplings are equivalent, you should construct contingency tables for each habitat (See Table \@ref(tab:contingency) for an example). 
Restrict each table to the five most common species in each habitat.  
Contingency tables can be constructed with pivot tables in Excel or the table() function in R.
Use chi-square tests to see if the proportions of species are different between canopy and saplings for each habitat.

```{r contingency, echo=FALSE, message=FALSE}
library(dplyr)
cont_dat = readr::read_csv("data/contingency_tbl_ex.r") %>% rename(` ` = "Tree_type")

knitr::kable(cont_dat, caption = "Abundance of most common canopy and sapling trees in the BFL old pasture habitat, Fall 2018.")
```


## Discussion

Some suggestions for discussion topics:

  - How could differences in sapling/canopy relative abundances inform possible successional trends? Based on your results, what would you predict about future dominant species in the habitats?
  - Did you observe anything else about the ecology or natural history of these areas that may help account for your results (e.g., drought stress, dead trees, invasive species, disease, etc)? 
  - What may be driving the differences in the habitat types? Considering the history of BFL may be helpful in explaining some of this.
  - Develop a likely scenario for the past and future decades of tree population dynamics in the woodlands of BFL.
  - How has drought and oak wilt affected the tree community at BFL? How might a scenario for succession based on currently healthy trees be changed if we include the information on stressed/dead trees?
  - How could this line of research be expanded upon in future work?  
  
## General Comments (Post-Review)

### Figures

Review the guidelines in the Figures chapter. Specifically: 

  - Don't use figure titles; anything that could go there should be in the caption instead.
  - Use colors that will work in black and white, 
  - Number figures in the order they're cited in the main text; they should also be arranged in this order (e.g., figure 2 shouldn't be placed before figure 1). 
  - If the figure contains new data (which all of these should), it should be cited in the results. 
  - Figure numbers shouldn't have decimals in them (e.g., no Figures 1.1, 2.3, etc).  If you have a multi-panel figure, refer to specific panels as Figure 1a, 1b, etc.
  - If you have a multi-panel figure, there should be a single caption that explains what each of the panels are.
  - Don't use bar plots for group means; boxplots or violin plots are usually better. This is a very inefficient way to present 3 data points, and it provides no information about the data's variability. Boxplots or violin plots are better options. (Note that bar plots are still fine for counts or proportions).
  - No gridlines (see below)
  
To remove gridlines in excel, just click on them and delete them.  Base plots in R (with the `plot()` function) shouldn't have them by default.  To remove them with `ggplot2` figures, put this near the top of the code:

```{r, echo = TRUE, eval = FALSE}
install.packages("cowplot") # if you don't have this installed; run once
library(cowplot)
theme_set(theme_cowplot()) # this makes your ggplots look nicer until you restart R
# ggplot commands here
```

### Write this like you're trying to publish it

You should write these reports as if you were writing a manuscript for a journal. Pretend it's not a class paper; don't say "For this lab, our assignment was..." or "the other students..." Write like it's a research project; you came up with the hypotheses and methods and the other students in the class are your collaborators.  

When describing the data collection, you need to describe how all of this year's data was collected, for all groups.  Instead of saying "we sampled three points in per habitat," say "five groups each sampled three points per habitat." As part of this, don't talk about combining your group's data with the rest.

Finally, you need to have a real title. 

### Describe the habitats 

Since the different habitat types are an important part of this paper, you should describe them in a reasonable amount of detail; this could go in the intro or methods section, depending on how you wrote it.  

### The Analysis

Formally, the chi-squared test evaluates whether the rows and columns of a contingency table are independent of each other.  In the context of this project, that's equivalent to testing whether the Canopy:Sapling ratio was equal for each species OR if the prevalence of each species was the same for canopy and sapling trees.  

Several people misinterpreted the analysis as examining if there were more canopy or sapling trees. 
This wouldn't work, because the chi-square test cannot tell you anything about the actual number of trees and because the way you collected the data (a fixed number of trees per point) was not set up to answer this question.

The Chi-square analysis was often described incorrectly in the methods; an unfamiliar reader would not know what you were testing. When you're describing the test, you don't necessarily need to state the null/alternative hypotheses, but you need to make it clear what question you're answering with it.

It's also worth remembering that the data we collect is generally not incredibly high precision. Our measurements generally don't have the ability to distinguish between a mean density of 437.11 and 437.14; depending on the sample size, we may not even be able to distinguish between 437 and 442. Being overly precise doesn't help.  If your p-values are less than 0.0001, just report "p < 0.0001."


### Keeping things in the right sections

Many people restated the methods used for their statistical analyses in the results. For example: 

    "I ran chi-squared tests, and they showed significant differences in the old quarry (chi-square = ..., df = ..., p = ...), the old pasture (...) and the river terrace (...)." 

Don't do this; instead, only provide the result. A better re-work of the above sentence would be:

    "Species composition was significantly different between canopy and sapling trees in the old quarry (...), the old pasture (...), and the river terrace (...)."

Don't put anything in the intro or methods that is actually a result.  Many people listed the most common species found in each habitat in their "study area" sections. Since this was a major part of this lab, listing the dominant species should be accompanied by a citation to show evidence that this is already known. In general, if you find yourself saying that the most common species "were" something, then it sounds like you are talking about your own observations; if you say that they "are" or "have been" something, then it sounds like you're talking about more general patterns.  

### Basic style and format rules

  - Don't capitalize things that don't need capitalization (e.g., the point-quarter technique).
  - Most acronyms should be defined before their first use (though there are a few exceptions, like ANOVA).
  - Don't say (p-value = x), say (p = x).
  - If a paper has more than two authors, cite it as "(Smith et al., 2009)," not "(Smith, Johnson, Franks, and Brown, 2009).  

### Other Comments

  - The "fisher test" is properly called Fisher's exact test (with "Fisher" capitalized).
  - If you used a stats program for a bunch of different things (as you usually do w/ R or Excel), mention it at the end of the section, not the beginning.
  - RStudio is an interface for doing analysis with R; if you used it, you should say you did your analysis in R.
  - Make sure to explain that sample points were selected differently in pond/old pasture than in other habitats. 
  - In the discussion, if you are listing several possible interpretations of your results, it's a good idea to put the most interesting one(s) ahead of the "this is all random noise" option.  
